A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-10 main.tex
4.4 note
Edits to 3.6

@@

Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\bX \boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align*}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.

This result also allows us to mathematically justify the statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure. We can quantify the proportion of the environmental effect which is in the column space of $\boldsymbol{P}_{\bX}$ using the ratio 
\begin{equation}
    \label{eqn:ratio}
    \anna{||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2 \in [0, 1]}.
\end{equation} 
In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. Thus the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates.

In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure \anna{and has the potential to bias SNP-phenotype associations can be inferred from $\bX$ and quantified by \anna{\eqref{eqn:ratio}}. We will use this metric in our analysis of our simulation results in Section \ref{sec:results}. Given the high-dimensional nature of our data, we use a low-rank approximation of $(\bX^\T \bX)^{-1}$. We use this approach as opposed to a generalized inverse in order to ensure a unique solution.}

%------------------------------%
\subsection{Simulation study}
%------------------------------%
\subsubsection{Pseudophenotype generation}

@@

\subsubsection{Design matrix generation}
We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data PATRICK TO DO}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Semi-Empirical data, respectively. $p = 1000$ for all data types, $n = \{200, 400, 600, 800 \}$ for the 1D Linear Admixture and Independent Subpopulations data, and 50 causal SNPs were used for all data types.

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered: (1) fine and (2) coarse. We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, or founding populations, which corresponds to a finer grained continuum of environmental heterogeneity. Whether fine or coarse, the simulated data were generated such that subpopulations, or founding populations, are of equal size. We note that the fine and coarse subpopulation structures are distinct from the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data. The Semi-Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

\subsubsection{Structure of environmental effects}
For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ structures were considered in order to investigate the effects of various forms of environmental confounding.

 Our initial hypothesis was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that are nonlinear in nature and as such do not lend themselves to such an approximation. To that end we investigated two forms of environmental effect magnitudes: (1) exponential, and (2) dichotomous. In addition to the magnitudes of effects, we also considered whether individuals in genetically similar subpopulations would experience similar environmental effects. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

 To evaluate the performance of penalized LMM methods in both scenarios, we consider exponential and dichotomous effect magnitudes with are concordant or discordant in nature, for a total of four distinct $\bgamma$ structures. We define concordant effects as those that are similar among genetically similar subpopulations, and discordant effects as those that are not. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient structures. 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure. 
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{$\bgamma$ structures}
    \label{fig:gamma_structures}
\end{figure}

\subsubsection{$\lambda$ selection}
The $\lambda$ that allowed for the first 50 variables to enter the model was selected for each method, where 50 was the true number of causal SNPs. This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. Alternative $\lambda$ selection methods were also considered. Cross-validation has been found to be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

\anna{re-create niter\_parLambda fig focusing on the eta = xi = 0.8 seting and displaying all data points - not just averages}

>>>>

\pb{How about ``Partitioning the effects of environmental confounding''}

@@

Although we derive these quantities under model \eqref{eqn:lmm_reg}, where $\bZ = \bX$, these results also hold for the more general model \eqref{eqn:model}, where $\bZ \ne \bX$. We see this by decomposing $\bZ \bgamma$ into components in and orthogonal to the column space of the projection matrix of $\bX$.  Let $\boldsymbol{P}_{\bX} = \bX (\bX^\T \bX)^{-1} \bX^\T$ denote the projection matrix of $\bX$, with $\mathcal{C}(\boldsymbol{P}_{\bX})$ and $\mathcal{N}(\boldsymbol{P}_{\bX})$ denoting the column and null spaces of $\boldsymbol{P}_{\bX}$, respectively. Then there exist $\boldsymbol{\tau}, \boldsymbol{\psi}: \bZ \bgamma = \bX \boldsymbol{\tau + \psi}$, where $\bX \boldsymbol{\tau} \in \mathcal{C}(\boldsymbol{P}_{\bX})$ and $\boldsymbol{\psi} \in \mathcal{N}(\boldsymbol{P}_{\bX})$.  This in turn means that
\begin{align*}
    \by &= \bX\bbeta + \bZ\bgamma + \beps\\
    &= \bX\bbeta + \bX \boldsymbol{\tau} + \boldsymbol{\psi} + \beps\\
    &=  \bX \bbeta{^*} + \beps{^*};
\end{align*}
note that $\boldsymbol{\tau}$ quantifies bias in the estimated $\bbetaHat^*$ and $\boldsymbol{\psi}$ is absorbed by the residual error term. In other words, any model specified in terms of $\bZ$ and $\bgamma$ is equivalent to a parameterization using $\bX$ and $\boldsymbol{\tau}$.  This result provides the explicit mathematical justification for our statement from section \ref{sec:background}, that SNP-phenotype associations will be unbiased in the presence of environmental heterogeneity which does not correspond to population structure.

Furthermore, the proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ (i.e., the portion that will cause confounding) is given by 
\begin{equation}
    \label{eqn:ratio}
    \anna{||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2},
\end{equation}
\pb{should we have a symbol for this?}
which can range from 0 (no confounding) to 1 (entirely confounding). In the case where $\bZ \bgamma$ is entirely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, but the residual variance will be increased. Although we focus our attention on accurately estimating fixed SNP effects, we note that this result could lead to substantial bias in narrow-sense heritability estimates. \pb{Hmm...could this sentence be moved somewhere else?  I feel that narrow-sense heritability deserves a paragraph somewhere}

\anna{In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure and has the potential to bias SNP-phenotype associations can be inferred from $\bX$ and quantified by \eqref{eqn:ratio}. We will use this metric in our analysis of our simulation results in Section \ref{sec:results}. Given the high-dimensional nature of our data, we use a low-rank approximation of $(\bX^\T \bX)^{-1}$. We use this approach as opposed to a generalized inverse in order to ensure a unique solution.}

  \pb{Need to mention that this ratio is difficult/impossible to estimate with real data, but can be calculated in simulations and is informative, etc.}
In practical applications $\bZ \bgamma$ is unlikely to be entirely orthogonal to population structure. However, the degree to which it reflects population structure \anna{and has the potential to bias SNP-phenotype associations can be inferred from $\bX$ and quantified by \anna{\eqref{eqn:ratio}}. We will use this metric in our analysis of our simulation results in Section \ref{sec:results}. Given the high-dimensional nature of our data, we will approximate $(\bX^\T \bX)^{-1}$ using a low-rank approximation. We use this approach as opposed to that of a generalized inverse in order to ensure a unique solution.}\pb{no need to mention generalized inverses}

%------------------------------%
\subsection{Simulation study}
%------------------------------%

\subsubsection{Generation of synthetic phenotypes}

\pb{I prefer ``synthetic'' phenotype to pseudophenotype}

@@

\subsubsection{Design matrix generation}

\pb{Let's not call it a ``design matrix''}

\pb{Also, we're generating $G$ matrices, correct?  And then obtaining $X$ from standardization?  Wouldn't hurt to remind people of that here}

\pb{Vanderwas citation: https://doi.org/10.1186/s12890-015-0040-0}

We generated these pseudophenotypes using three types of high-dimensional ($p > n$) $\bX$ matrices: (1) simulated SNP data from admixed subpopulations, (2) simulated SNP data from independent subpopulations, and (3) real genome-wide SNP data from known caucasian ($n = 417$), African American ($n = 134$), and Hispanic ($n = 37$) subpopulations \anna{cite Vanderwas data PATRICK TO DO}. We refer to these as the 1D Linear Admixture, Independent Subpopulations, and Semi-Empirical data, respectively. \pb{Don't start sentences with mathematical symbols} $p = 1000$ for all data types, $n = \{200, 400, 600, 800 \}$ for the 1D Linear Admixture and Independent Subpopulations data, and 50 causal SNPs were used for all data types.

\pb{Can't we just call the last one ``Empirical''?}

\subsubsection{Subpopulation structure}
For the fully simulated data, two subpopulation structures were considered, which we refer to as ``fine'' and ``coarse''.  We define a coarse structure to be one characterized by a few large subpopulations intended to model large-scale environmental heterogeneity and population stratification, such as that described in our motivating air quaility example. Conversely, we define a fine subpopulation structure to be one characterized by many small subpopulations, representing a more subtle continuum of environmental heterogeneity \pb{do you know, are the subpopulations more similar in the fine case?}. Whether fine or coarse, the simulated data were generated such that subpopulations (also known as founding populations) are of equal size. We note that the distinction between fine and coarse structure is an entirely different concept than the level of admixture in the simulated data. For example, the discrete nonoverlapping subpopulations in the simulated Independent Subpopulations data result in an RRM which more directly corresponds to the discrete and fixed environmental effects, fine or coarse, than do the continuous and overlapping subpopulations in the simulated 1D Linear Admixture data \pb{?????}. The Semi-Empirical data is comprised of three racial subpopulations; one relatively large, one small, and one in the middle. Thus it represents a hybrid of our simulated coarse and fine structures, with little apparent admixture. Supplementary Figures \anna{?, ?, and ? MAKE THESE AND MAKE THEM PRETTY} illustrate the simulated kinship structures in the fine and coarse subpopulation settings, as well as that present in the empirical SNP data.

\subsubsection{Structure of environmental effects}

\pb{Perhaps ``Patterns of environmental effects'' would be better?}

For each $\bX$ matrix, $\bZ$ was a corresponding incidence matrix of 1s and 0s identifying subpopulation, or founding population membership \pb{you don't need to keep saying ``or founding population'' over and over again}, which assigns the corresponding environmental effect found in $\bgamma$. Various $\bgamma$ structures were considered in order to investigate the effects of various forms of environmental confounding.

Our initial hypothesis was that structures of $\bgamma$ that can be well-approximated by a linear combination of the observed genetic data would be well-approximated and corrected for by PC-Lasso, whereas LMM-Lasso may be more able to correct for subpopulation-specific effects that are nonlinear in nature and as such do not lend themselves to such an approximation \pb{is the ``linearity'' the key issue here?  I'd say it's more a matter of complexity: PC-Lasso would work for simple patterns, but can't represent more complex patterns}. To that end we investigated two forms of environmental effect magnitudes: (1) exponential, and (2) dichotomous \pb{neither of which are linear, so this doesn't transition well}. In addition to the magnitudes of effects, we also considered whether individuals in genetically similar subpopulations would experience similar environmental effects \pb{Shouldn't we lead with this?  I mean, isn't this the most important concept in this section?}. For example, a geographic environmental exposure such as the air quality described in our motivating example is likely to be similar among genetically related individuals due to concordance between genetic clusters and geography. However, behavior-based environmental exposures may be less likely to be shared by subpopulations which are genetically and geographically similar. As an example, \citet{ng2014smoking} found considerable within-region variation in smoking rates in Africa, Asia, and South America.

 To evaluate the performance of penalized LMM methods in both scenarios, we consider exponential and dichotomous effect magnitudes with are concordant or discordant in nature, for a total of four distinct $\bgamma$ structures. We define concordant environmental effects as those that are similar among genetically similar subpopulations, and discordant environmental effects as those that are not. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient structures. 1000 simulations were performed for each unique combination of parameters, datatype, subpopulation structure, and environmental effect structure. \pb{need to describe somehow that subpopulations are ordered with respect to how similar they are?}
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{$\bgamma$ structures \pb{probably want a better caption}}
    \label{fig:gamma_structures}
\end{figure}

\subsubsection{Tuning parameter selection}
Selection of the tuning parameter $\lambda$ plays an important role in all penalized regression methods, with many possible approaches.  Here, for each model, we selected the $\lam$ value such that 50 SNPs were selected (recall that 50 is also the true number of causal SNPs). This method for selecting $\lambda$ was chosen in an attempt to compare the performance of all methods fairly, and as a strategy to evaluate methods in terms of variable selection, as it allows for direct comparison of the first 50 variables each method lets into the model. Alternative $\lambda$ selection methods were also considered. \pb{This seems way too strong}  Cross-validation has been found to be inadvisable for LMMs, and in the presence of non-inedpendent samples as is the case here. While information criterion methods such as BIC present another option for selecting $\lambda$, they may be unstable in high dimensions. In addition, because Lasso, PC-Lasso, and LMM-Lasso use different likelihoods, an information-criterion-based approach would not lend itself to comparable results across methods.

@@

\anna{See figure options: niter\_parLambda fig focusing on the eta = xi = 0.8 seting and displaying all data points - not just averages}
